---
title: "STING Pathway in Sjogren's Disease, figure 1 code"
author: "Joseph Twohig"
output:
  html_document:
    toc: true
    toc-location: left
  pdf_document:
    toc: true
---

```{r message=FALSE, warning=FALSE, results = FALSE}
library(readxl)
library(tidyverse) 
library(edgeR) # DGE testing
library(DESeq2) # DGE testing
library(clusterProfiler) # GO
library(AnnotationDbi) # GO
library(patchwork) # Plotting
library(gridExtra) # Plotting
library(ggrepel) # Plotting
library(GSVA) # Performing GSVA
library(msigdbr) # Pulling gene sets
library(ComplexHeatmap) # For heatmaps
library(org.Hs.eg.db) # Get gene mapping
library(circlize)  # For color functions
library(dendextend)  # For dendrogram manipulation
library(scales) # Color scales
library(ggpubr) # Stat tests

setwd("/data/Warner_Lab/bulk_sting_analysis")
```

Load in expression matrices and metadata

```{r}
metadata <- read.csv("metadata.csv")
metadata$SSA <- case_when(
  grepl("1", metadata$SSA) ~ "Positive",
  grepl("0", metadata$SSA) ~ "Negative",
  TRUE ~ NA_character_
)
metadata$SSB <- case_when(
  grepl("1", metadata$SSB) ~ "Positive",
  grepl("0", metadata$SSB) ~ "Negative",
  TRUE ~ NA_character_
)

gland <- read.delim("/data/Warner_Lab/STING_BulkRNA_SjD_HV_nonSjD/MSG_with_corrected_sample_names/report/RawCountFile_filtered.txt",row.names=1,check.names=FALSE)%>% 
  dplyr::select(metadata$MSG_Samples)

blood <- read.delim("/data/Warner_Lab/STING_BulkRNA_SjD_HV_nonSjD/Blood_with_corrected_sample_names/report/RawCountFile_filtered.txt",row.names=1,check.names=FALSE) %>%
  dplyr::select(metadata$Blood_Samples)
```

Convert the Ensembl IDs in the counts matrices to gene symbols

```{r message=FALSE, warning=FALSE}
# Make function to convert and filter for gene symbols
ens_to_symbol <- function(matrix) {
  ens <- rownames(matrix)
  symbols <- mapIds(org.Hs.eg.db, keys = ens,
                    column = c('SYMBOL'), keytype = 'ENSEMBL')
  symbols <- symbols[!is.na(symbols)]
  symbols <- symbols[!duplicated(symbols)]
  symbols <- symbols[match(rownames(matrix), names(symbols))]
  symbols <- symbols[!is.na(symbols)]
  matrix <- matrix[names(symbols), ]
  rownames(matrix) <- symbols
  return(matrix)
}

# Convert gland and blood count matrices
gland <- ens_to_symbol(gland)
blood <- ens_to_symbol(blood)
```

# 1b: Supervised, expression-based hierarchical clustering

Take the top variable genes and cluster samples based on expression. Then
map the dendrogram onto the metadata and visualize. 

```{r}
# Prepare metadata subset and factor levels
metadata_subset <- metadata %>%
  dplyr::select(Age, Sex, condition, SSA, SSB, Focus.Score, RF, ANA,
         RACE, Type.I.IFN.Bioassay.AVG, WUS, WSSF) %>%
  mutate(
    Focus.Score = case_when(
      Focus.Score == 0 ~ "None (0)",
      Focus.Score == 1 ~ "Low (1)",
      Focus.Score >= 2 & Focus.Score < 4 ~ "Medium (2-3)",
      Focus.Score >= 4 ~ "High (≥4)"
    ),
    Focus.Score = factor(Focus.Score, levels = c("None (0)", "Low (1)", "Medium (2-3)", "High (≥4)")),
    across(c(Sex, condition, SSA, SSB, RACE), as.factor)
  )

# Filter metadata to plot
# metadata_for_clustering <- metadata_subset %>% dplyr::select(-SPIT_number, -ESSDAI)

# Split metadata by condition
metadata_split <- split(metadata_subset, metadata_subset$condition)

# Normalize expression and select top 300 variable genes ---
norm_gland <- log2(gland + 1)
top_genes <- norm_gland %>%
  apply(1, var) %>%
  sort(decreasing = TRUE) %>%
  head(300) %>%
  names()
top_expr <- norm_gland[top_genes, ]

# Functions
create_color_scale <- function(values, colors) {
  values <- na.omit(values)
  if (length(unique(values)) > 1) {
    colorRamp2(c(min(values), max(values)), colors)
  } else if (length(unique(values)) == 1) {
    colorRamp2(c(values[1] - 0.1, values[1] + 0.1), colors)
  } else {
    colorRamp2(c(0, 1), colors)
  }
}

make_dendrogram <- function(condition) {
  samples <- metadata %>% filter(condition == !!condition) %>% pull(MSG_Samples)
  expr <- top_expr[, samples, drop = FALSE]
  hclust(dist(t(expr))) %>% as.dendrogram()
}

make_annotation <- function(meta) {
  HeatmapAnnotation(
    show_annotation_name = FALSE,
    Age = meta$Age,
    Sex = meta$Sex,
    Condition = meta$condition,
    FS = meta$Focus.Score,
    SSA = meta$SSA,
    SSB = meta$SSB,
    ESSDAI = meta$ESSDAI,
    WUS = meta$WUS,
    WSSF = meta$WSSF,
    RACE = meta$RACE,
    col = list(
      Age = create_color_scale(meta$Age, c("yellow", "orange")),
      Sex = c("FEMALE" = "pink", "MALE" = "lightblue"),
      Condition = c("SjD" = "red", "nonSjD" = "grey", "HV" = "blue"),
      FS = c("High (≥4)" = "#7f03fc", "Medium (2-3)" = "#524fff", "Low (1)" = "#7e9cf7", "None (0)" = "#b3cff5"),
      SSA = c("Positive" = "darkred", "Negative" = "white"),
      SSB = c("Positive" = "darkblue", "Negative" = "white"),
      ESSDAI = create_color_scale(meta$ESSDAI, c("green", "darkgreen")),
      WUS = create_color_scale(meta$WUS, c("white", "turquoise")),
      WSSF = create_color_scale(meta$WSSF, c("white", "springgreen")),
      RACE = c(
        "BLACK OR AFRICAN AMERICAN" = "purple",
        "ASIAN" = "orange",
        "WHITE" = "cyan",
        "MULTIPLE RACES" = "brown",
        "UNKNOWN" = "grey"
      )
    )
  )
}

make_heatmap <- function(n_col, dend, annotation) {
  Heatmap(
    matrix(nc = n_col, nr = 0),
    cluster_columns = dend,
    show_row_dend = FALSE,
    show_column_dend = TRUE,
    column_names_gp = gpar(fontsize = 10),
    top_annotation = annotation,
    show_heatmap_legend = FALSE
  )
}

# Create heatmaps
heatmap_SjD <- make_heatmap(nrow(metadata_split$SjD),make_dendrogram("SjD"),make_annotation(metadata_split$SjD))
heatmap_nonSjD <- make_heatmap(nrow(metadata_split$nonSjD),make_dendrogram("nonSjD"),make_annotation(metadata_split$nonSjD))
heatmap_HV <- make_heatmap(nrow(metadata_split$HV),make_dendrogram("HV"),make_annotation(metadata_split$HV))

combined_heatmap <- heatmap_SjD + heatmap_nonSjD + heatmap_HV

# Plot and save
draw(combined_heatmap, annotation_legend_side = "bottom")

# svg("figure1/metaHeatmapExpressionSupervised.svg", width = 12, height = 4)
# draw(combined_heatmap, annotation_legend_side = "bottom")
# dev.off()
# 
# png("figure1/metaHeatmapExpressionSupervised.png", width = 12, height = 4, units = "in", res = 300)
# draw(combined_heatmap, annotation_legend_side = "bottom")
# dev.off()
```


# 1c,d: Principal Component Analysis (PCA)

```{r message=FALSE, warning=FALSE}
# Plot function
pcaplot <- function(matrix, tissue_title, ellipse = TRUE, ellipse.prob = 0.8) {
  ddsHTSeq <- DESeqDataSetFromMatrix(countData = matrix, colData = metadata['condition'], design = ~condition)
  dds <- DESeq(ddsHTSeq)
  rlt <- varianceStabilizingTransformation(dds, blind = FALSE)
  vsd_matrix <- assay(rlt)
  # Select variable genes
  top500_genes <- head(order(rowVars(vsd_matrix), decreasing = TRUE), 500)
  top500_matrix <- vsd_matrix[top500_genes, ]
  pca_results <- prcomp(t(top500_matrix), scale. = TRUE)
  percentVar <- pca_results$sdev^2 / sum(pca_results$sdev^2)
  pca_data <- as.data.frame(pca_results$x)
  pca_data$condition <- metadata$condition
  g <- ggplot(pca_data, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3) +
    scale_color_manual(values = c("SjD" = "red", "nonSjD" = "grey", "HV" = "blue")) +
    labs(
      title = paste0("PCA of ", tissue_title),
      x = paste0("PC1, Variance Explained: ", round(percentVar[1] * 100, digits = 2), "%"),
      y = paste0("PC2, Variance Explained: ", round(percentVar[2] * 100, digits = 2), "%")
    ) +
    theme_minimal() +
    theme(
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 10),
      plot.title = element_text(size = 12),
      legend.position = "right"
    )
  # Add ellipse by condition
  if (ellipse) {
    g <- g + stat_ellipse(aes(fill = condition), alpha = 0.2, level = ellipse.prob, geom = "polygon") +
      scale_fill_manual(values = c("SjD" = "red", "nonSjD" = "grey", "HV" = "blue"))
  }
  return(g)
}

msgpca <- pcaplot(gland, "Minor Salivary Gland Samples")
msgpca

# ggsave(file="figure1/glandPca.svg", plot=msgpca, width=5, height=3)

bloodpca <- pcaplot(blood, "Blood Samples")
bloodpca

# ggsave(file="figure1/bloodPca.svg", plot=bloodpca, width=5, height=3)
```

# 1e: Gene Ontology (GO) analysis

```{r message=FALSE, warning=FALSE}
# Reformat metadata to make single comparison
metadata$disease <- case_when(
  grepl("nonSjD", metadata$condition) ~ "nonSjD/HV",
  grepl("HV", metadata$condition) ~ "nonSjD/HV",
  TRUE ~ "SjD"
)

condition <- factor(metadata$disease[match(colnames(gland), metadata$MSG_Samples)])
coldata <- data.frame(row.names = colnames(gland), condition)

# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = gland, colData = coldata, design = ~condition)
dds <- DESeq(dds)

# Get differential expression results (SjD vs nonSjD/HV)
res <- results(dds, contrast = c("condition", "SjD", "nonSjD/HV"))

# Filter significant DEGs
sigs <- na.omit(res)
sigs <- sigs[sigs$padj < 0.05 & sigs$baseMean > 50, ]

# Select upregulated genes (log2FC > 0.5)
genes_to_test <- rownames(sigs[sigs$log2FoldChange > 0.5, ])

# Perform GO enrichment analysis (Biological Process)
GO_results <- enrichGO(gene = genes_to_test, OrgDb = org.Hs.eg.db, keyType = "SYMBOL", ont = "BP")

# Generate barplot
plot <- barplot(GO_results, showCategory = 8, main = "GO Enrichment MSG SjD vs nonSjD/HV")
plot

# ggsave(file="figure1/glandGObarplot.svg", plot=plot, width=8, height=8)
```

# 1f: Gene Set Variation Analysis (GSVA)

```{r message=FALSE, warning=FALSE}
# Retrieve pathway data from msigdbr, filtering relevant pathways
pathways <- rbind(msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY") %>% 
  filter(gs_name %in% c("KEGG_CYTOSOLIC_DNA_SENSING_PATHWAY", "KEGG_RIG_I_LIKE_RECEPTOR_SIGNALING_PATHWAY")))

# Create gene set for GSVA input
gene_sets <- split(pathways,
                   pathways$gs_name) %>% 
  lapply(`[[`,
         "gene_symbol")
gene_sets$IFN_Type_I_28genes <- c("CXCL10","DDX60","EPSTI1","GBP1","HERC5",
                        "HERC6","IFI27","IFI44","IFI44L","IFI6",
                        "IFIT1","IFIT2","IFIT3","IFIT5","ISG15",
                        "LAMP3","LY6E","MX1","OAS1","OAS2",
                        "OAS3","OASL","RSAD2","RTP4","SIGLEC1",
                        "SOCS1","SPATS2L","USP18")
gene_sets$IFN_Type_II_8genes <- c("FCGR1A","CXCL9","IFI30","CIITA","SERPING1",
                        "IRF1","CXCL10","GBP8")

# Normalize counts data matrices before running GSVA
norm_gland <- cpm(gland, log = T)
gsva_gland <- gsva(gsvaParam(exprData = norm_gland, geneSets = gene_sets))
gland_heat <- gsva_gland
rownames(gland_heat) <- c("DNA Sensing", "RNA Sensing","28 gene Type I IFN", "8 gene Type II IFN")

# Creating a dataframe with GSVA results and metadata to use in visualization
targets <- metadata %>% 
  dplyr::select("SPIT_number", "condition")

# Heatmap annotation
col_condition <- c("SjD" = "red", "nonSjD" = "grey", "HV" = "blue")

# Construct heatmap
glandGSVAHeatmap <- Heatmap(
  gland_heat,
  column_title = "Minor Salivary Gland GSVA Heatmap",
  name = "Expression",
  show_row_names = TRUE,
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 10),
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 12),
    at = seq(-2, 2, length.out = 5)
  ),
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  top_annotation = HeatmapAnnotation(
    Condition = targets$condition,
    col = list(
      Condition = col_condition
    ),
    annotation_legend_param = list(
      title_gp = gpar(fontsize = 12)
    )
  )
)

glandGSVAHeatmap

# svg(file="figure1/glandGSVAHeatmap.svg", width=8, height=5.5)
# draw(glandGSVAHeatmap)
# dev.off()
# png(file="figure1/glandGSVAHeatmap.png", width=7, height=5.5, units = "in", res = 500)
# draw(glandGSVAHeatmap)
# dev.off()
```